---
title: "Effect of Citronellal Odor Puffs on Cockroach Neurons 1, 2, and 3"
format: html
editor: visual
---

## Goal

We compare the effect of citronellal odor puff stimulation on the firing rate of two cockroach antennal lobe neurons (Neuron 1, Neuron 2 and Neuron 3).

We will:

1.  Visualize spike trains.
2.  Form a qualitative hypothesis.
3.  Fit one NHPP model per neuron.
4.  Visualize stimulus effects.
5.  Determine whether the effect differs across neurons.

------------------------------------------------------------------------

```{r setup, include=FALSE}
library(mgcv)

# load data
download.file(
  url = 'https://ruizt.quarto.pub/stat545/data/cockroach.RData',
  destfile = 'cockroach.RData'
)
load('cockroach.RData')

# panel raster plot from ev_nX (list of spike-time vectors), with stimulus window highlighted
plot_rasters_from_ev <- function(ev, n_trials = 10, trials = NULL,
                                 stim_on = 5.99, stim_off = 6.49,
                                 ncol = 5, T_end = 15, seed = 1) {
  stopifnot(is.list(ev))
  all_trials <- names(ev)

  if (is.null(trials)) {
    set.seed(seed)
    trials <- sample(all_trials, size = min(n_trials, length(all_trials)))
  } else {
    trials <- intersect(trials, all_trials)
  }

  n_trials <- length(trials)
  nrow <- ceiling(n_trials / ncol)

  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  par(mfrow = c(nrow, ncol), mar = c(2.2, 2.2, 2.0, 0.5), mgp = c(1.5, 0.5, 0))

  for (tr in trials) {
    times <- ev[[tr]]
    times <- times[is.finite(times) & times >= 0 & times <= T_end]

    plot(NA, xlim = c(0, T_end), ylim = c(0, 1),
         xlab = "", ylab = "", yaxt = "n", main = tr)

    rect(stim_on, 0, stim_off, 1, border = NA, col = adjustcolor("red", 0.5))

    if (length(times)) segments(times, 0.2, times, 0.8, lwd = 1)

    box()
    axis(1, at = 5.99, labels = 5.99, cex.axis = 0.75)
  }
}
```

## Raster Plots: Neuron 1

If stimulation affects firing rate, we expect spike density to change during the red stimulus window (5.99–6.49 s).

```{r}
plot_rasters_from_ev(ev_n1, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 1):**

Based on the raster plots for Neuron 1, stimulation seems to lead to higher firing rate and dense spikes. This is shown by dark bands present in and following the stimulus, which is shown by the red area. We hypothesize that the odor puff simulation leads to more activity, in other words higher firing rate, in neuron 1.

------------------------------------------------------------------------

## Raster Plots: Neuron 2

If stimulation affects firing rate, we expect spike density to change during the red stimulus window (5.99–6.49 s).

```{r}
plot_rasters_from_ev(ev_n2, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 2):**

For neuron 2, there is some activity clustering during and following the stimulus. Similar to neuron 1, there are clear dark bands of high activity and increased firing rate after the stimulus in neuron 2. We hypothesize that the stimulus decreases firing rate and less activity in neuron 2.

------------------------------------------------------------------------

## Raster Plots: Neuron 3

```{r}
plot_rasters_from_ev(ev_n3, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 3):**

Unlike neurons 1 and 2, which exhibited more activity following the stimulus, neuron 3 exhibits gaps in activity after the stimulus. This suggests that the stimulus leads to a lower firing rate. So the logical hypothesis is that the stimulus leads to low activity in neuron 3.

------------------------------------------------------------------------

## Model Specification

We approximate the non-homogeneous Poisson process by binning time.

Assume:

$$
Y_{it} \sim \text{Poisson}(\lambda_i(t)\Delta t)
$$

Using the log link:

$$
\log(\lambda_i(t)) = \beta_0 + \beta_1 \cdot \text{stim}(t) + f(t) + b_{\text{trial}}
$$

Where:

-   $\beta_1$ = stimulation effect (log scale)
-   $f(t)$ = smooth baseline firing pattern
-   $b_{\text{trial}} \sim N(0,\sigma^2)$ = trial random effect
-   $\log(\Delta t)$ = offset

------------------------------------------------------------------------

## Fit Model: Neuron 1

```{r}
df_n1$trial <- factor(df_n1$trial)

fit_n1 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n1,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 1

-   The plot shows the **estimated smooth time effect** from the fitted GAM for neuron 1.
-   Specifically, we plot $\exp(\hat{f}(t))$, the **exponentiated smooth function of time**.
-   $f(t)$ captures **time-varying structure in the log firing rate** not explained by the stimulus term.
-   Exponentiating converts the smooth from the **log scale to a multiplicative scale** for firing intensity.
-   The **black curve** shows the estimated effect $\exp(\hat{f}(t))$.
-   The **dashed bands** represent an approximate **95% confidence interval**.
-   The **red shaded region** marks the **stimulus window (odor puff)**.
-   The **horizontal line at 1** indicates the **reference level of the smooth** (no multiplicative change).

```{r}
p1 <- plot(fit_n1, select = 1)[[1]]
x1 <- p1$x
y1 <- exp(p1$fit)
lo1 <- exp(p1$fit - 2 * p1$se)
hi1 <- exp(p1$fit + 2 * p1$se)

plot(x1, y1, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 1: Effect of Stimulation")

polygon(c(x1, rev(x1)), c(lo1, rev(hi1)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x1, y1, lwd = 2)
abline(h = 1, lty = 2)
```

Interpretation:

-   If curve \> 1 → increased firing
-   If curve \< 1 → suppression
-   If CI includes 1 → weak/no evidence

Because the curve and confidence bands lie entirely above 1 during and immediately following the stimulus, we have evidence that the stimulus leads to increased firing in neuron 1.

------------------------------------------------------------------------

## Fit Model: Neuron 2

```{r}
df_n2$trial <- factor(df_n2$trial)

fit_n2 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n2,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 2

```{r}
p2 <- plot(fit_n2, select = 1)[[1]]
x2 <- p2$x
y2 <- exp(p2$fit)
lo2 <- exp(p2$fit - 2 * p2$se)
hi2 <- exp(p2$fit + 2 * p2$se)

plot(x2, y2, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 2: Effect of Stimulation")

polygon(c(x2, rev(x2)), c(lo2, rev(hi2)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x2, y2, lwd = 2)
abline(h = 1, lty = 2)
```

Interpretation:

-   If curve \> 1 → increased firing
-   If curve \< 1 → suppression
-   If CI includes 1 → weak/no evidence

In neuron 2, like neuron 1, the curve and confidence bands lie entirely above 1 during and immediately following the stimulus. Therefore, we also have evidence that the stimulus leads to increased firing in neuron 2.

------------------------------------------------------------------------

## Fit Model: Neuron 3

```{r}
df_n3$trial <- factor(df_n3$trial)

fit_n3 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n3,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 3

```{r}
p3 <- plot(fit_n3, select = 1)[[1]]
x3 <- p3$x
y3 <- exp(p3$fit)
lo3 <- exp(p3$fit - 2 * p3$se)
hi3 <- exp(p3$fit + 2 * p3$se)

plot(x3, y3, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 3: Effect of Stimulation")

polygon(c(x3, rev(x3)), c(lo3, rev(hi3)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x3, y3, lwd = 2)
abline(h = 1, lty = 2)
```

In neuron 3, the curve and confidence bands lie entirely below 1 during and immediately following the stimulus. So for neuron 3, we have evidence that the stimulus leads to decreased firing.

------------------------------------------------------------------------

## **Final Comparison**

### **1. Do citronellal odor puffs affect firing rate?**

Yes. The models show that the multiplicative effect of the stimulus differs from 1 for all neurons.

-   **Neuron 1:** The multiplicative effect is greater than 1 during and shortly after the stimulus window. During this period, the confidence band lies **above** 1, indicating that the citronellal odor puff increases firing rate of this neuron.

-   **Neuron 2:** The multiplicative effect rises above 1 during the stimulus window, indicating that the citronellal odor puff increases firing rate while the stimulus is active. The confidence band does not include 1, so the increase is further supported. Immediately after the stimulus, both the multiplicative effect and confidence band drop below 1, indicating a decrease in firing rate following the stimulus.

-   **Neuron 3:** The multiplicative effect is **below** 1duringthe stimulus window, indicating that the citronellal odor puff suppressed the neuron firing rate while it was in effect. Immediately after the stimulus, however, the effect **rises**above 1, suggesting a delayed increase in firing rate. The confidence band lies below 1 during the stimulus window, further providing evidence that the odor puff suppressed the firing rate.

### **2. Is the effect uniform across neurons?**

Possible outcomes:

-   Both increase → uniform excitatory effect

-   One increases, one decreases → neuron-specific response

-   One significant, one null → heterogeneous sensitivity

-   Both null → no detectable effect

No, the effect of citronellal odor stimulation is **not** uniform across the neurons.

The direction and magnitude of the response differ across the three neurons.

-   **Neuron 1** shows a clear increase in firing rate during the stimulus window, and the confidence band lies above 1, indicating strong evidence of an excitatory effect.

-   **Neuron 2** also shows an increase during the stimulus, although the magnitude of the effect is smaller and the confidence band reflects greater uncertainty.

-   **Neuron 3**, in contrast, shows a decrease in firing rate during the stimulus window, with the multiplicative effect falling below 1 and the confidence band supporting the evidence of suppression.

Because two neurons show increased firing rate while one shows decreased firing rate, the response to the citronellal stimulation appears to be neuron-specific rather than uniform across neurons.

------------------------------------------------------------------------

## **Conclusion**

Based on the fitted NHPP models:

-   **Neuron 2:** increases during stimulus, decreases after

-   **Neuron 3:** decrease during stimulus, increase after

Therefore, citronellal odor puffs **do** affect firing rate, and the effect **is not** uniform across neurons.
