---
title: "Effect of Citronellal Odor Puffs on Cockroach Neurons 1, 2, and 3"
format: html
editor: visual
---

## Goal

We compare the effect of citronellal odor puff stimulation on the firing rate of two cockroach antennal lobe neurons (Neuron 1, Neuron 2 and Neuron 3).

We will:

1.  Visualize spike trains.
2.  Form a qualitative hypothesis.
3.  Fit one NHPP model per neuron.
4.  Visualize stimulus effects.
5.  Determine whether the effect differs across neurons.

------------------------------------------------------------------------

```{r setup, include=FALSE}
library(mgcv)

# load data
download.file(
  url = 'https://ruizt.quarto.pub/stat545/data/cockroach.RData',
  destfile = 'cockroach.RData'
)
load('cockroach.RData')

# panel raster plot from ev_nX (list of spike-time vectors), with stimulus window highlighted
plot_rasters_from_ev <- function(ev, n_trials = 10, trials = NULL,
                                 stim_on = 5.99, stim_off = 6.49,
                                 ncol = 5, T_end = 15, seed = 1) {
  stopifnot(is.list(ev))
  all_trials <- names(ev)

  if (is.null(trials)) {
    set.seed(seed)
    trials <- sample(all_trials, size = min(n_trials, length(all_trials)))
  } else {
    trials <- intersect(trials, all_trials)
  }

  n_trials <- length(trials)
  nrow <- ceiling(n_trials / ncol)

  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  par(mfrow = c(nrow, ncol), mar = c(2.2, 2.2, 2.0, 0.5), mgp = c(1.5, 0.5, 0))

  for (tr in trials) {
    times <- ev[[tr]]
    times <- times[is.finite(times) & times >= 0 & times <= T_end]

    plot(NA, xlim = c(0, T_end), ylim = c(0, 1),
         xlab = "", ylab = "", yaxt = "n", main = tr)

    rect(stim_on, 0, stim_off, 1, border = NA, col = adjustcolor("red", 0.5))

    if (length(times)) segments(times, 0.2, times, 0.8, lwd = 1)

    box()
    axis(1, at = 5.99, labels = 5.99, cex.axis = 0.75)
  }
}
```

## Raster Plots: Neuron 1

If stimulation affects firing rate, we expect spike density to change during the red stimulus window (5.99–6.49 s).

```{r}
plot_rasters_from_ev(ev_n1, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 1):**

Visually inspect whether spikes cluster inside the red region.

------------------------------------------------------------------------

## Raster Plots: Neuron 2

If stimulation affects firing rate, we expect spike density to change during the red stimulus window (5.99–6.49 s).

```{r}
plot_rasters_from_ev(ev_n2, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 2):**

Visually inspect whether spikes cluster inside the red region.

------------------------------------------------------------------------

## Raster Plots: Neuron 3

```{r}
plot_rasters_from_ev(ev_n3, n_trials = 6, ncol = 2)
```

**Hypothesis (Neuron 3):**

Compare spike density during stimulation to baseline and compare to Neuron 2.

------------------------------------------------------------------------

## Model Specification

We approximate the non-homogeneous Poisson process by binning time.

Assume:

$$
Y_{it} \sim \text{Poisson}(\lambda_i(t)\Delta t)
$$

Using the log link:

$$
\log(\lambda_i(t)) = \beta_0 + \beta_1 \cdot \text{stim}(t) + f(t) + b_{\text{trial}}
$$

Where:

-   $\beta_1$ = stimulation effect (log scale)
-   $f(t)$ = smooth baseline firing pattern
-   $b_{\text{trial}} \sim N(0,\sigma^2)$ = trial random effect
-   $\log(\Delta t)$ = offset

------------------------------------------------------------------------

## Fit Model: Neuron 1

```{r}
df_n1$trial <- factor(df_n1$trial)

fit_n1 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n1,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 1

We exponentiate the smooth to obtain multiplicative change in firing intensity:

$$
\text{Intensity Ratio} = \exp(\beta_1)
$$

```{r}
p1 <- plot(fit_n1, select = 1)[[1]]
x1 <- p1$x
y1 <- exp(p1$fit)
lo1 <- exp(p1$fit - 2 * p1$se)
hi1 <- exp(p1$fit + 2 * p1$se)

plot(x1, y1, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 1: Effect of Stimulation")

polygon(c(x1, rev(x1)), c(lo1, rev(hi1)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x1, y1, lwd = 2)
abline(h = 1, lty = 2)
```

Interpretation:

-   If curve \> 1 → increased firing
-   If curve \< 1 → suppression
-   If CI includes 1 → weak/no evidence

------------------------------------------------------------------------

## Fit Model: Neuron 2

```{r}
df_n2$trial <- factor(df_n2$trial)

fit_n2 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n2,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 2

We exponentiate the smooth to obtain multiplicative change in firing intensity:

$$
\text{Intensity Ratio} = \exp(\beta_1)
$$

```{r}
p2 <- plot(fit_n2, select = 1)[[1]]
x2 <- p2$x
y2 <- exp(p2$fit)
lo2 <- exp(p2$fit - 2 * p2$se)
hi2 <- exp(p2$fit + 2 * p2$se)

plot(x2, y2, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 2: Effect of Stimulation")

polygon(c(x2, rev(x2)), c(lo2, rev(hi2)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x2, y2, lwd = 2)
abline(h = 1, lty = 2)
```

Interpretation:

-   If curve \> 1 → increased firing
-   If curve \< 1 → suppression
-   If CI includes 1 → weak/no evidence

------------------------------------------------------------------------

## Fit Model: Neuron 3

```{r}
df_n3$trial <- factor(df_n3$trial)

fit_n3 <- gam(
  y ~ stim + s(t, bs = "cr", k = 15) + s(trial, bs = "re"),
  family = poisson(),
  offset = log_dt,
  gamma = 1.4,
  data = df_n3,
  method = "REML"
)
```

------------------------------------------------------------------------

## Stimulus Effect on Intensity: Neuron 3

```{r}
p3 <- plot(fit_n3, select = 1)[[1]]
x3 <- p3$x
y3 <- exp(p3$fit)
lo3 <- exp(p3$fit - 2 * p3$se)
hi3 <- exp(p3$fit + 2 * p3$se)

plot(x3, y3, type = "n",
     xlab = "t (s)",
     ylab = "multiplicative effect",
     main = "Neuron 3: Effect of Stimulation")

polygon(c(x3, rev(x3)), c(lo3, rev(hi3)),
        border = NA,
        col = adjustcolor("gray", 0.25))

usr <- par("usr")
rect(5.99, usr[3], 6.49, usr[4],
     border = NA,
     col = adjustcolor("red", 0.20))

lines(x3, y3, lwd = 2)
abline(h = 1, lty = 2)
```

------------------------------------------------------------------------

## Final Comparison

Answer the key questions:

### 1. Do citronellal odor puffs affect firing rate?

-   Does the multiplicative effect differ from 1?
-   Does the confidence band exclude 1?

### 2. Is the effect uniform across neurons?

Compare:

-   Magnitude
-   Direction
-   Uncertainty

Possible outcomes:

-   Both increase → uniform excitatory effect
-   One increases, one decreases → neuron-specific response
-   One significant, one null → heterogeneous sensitivity
-   Both null → no detectable effect

------------------------------------------------------------------------

## Conclusion

Based on the fitted NHPP models:

-   Neuron 2: (describe result)
-   Neuron 3: (describe result)

Therefore, citronellal odor puffs **\[do / do not\]** affect firing rate, and the effect **\[is / is not\]** uniform across neurons.
